---
title: "Pittsburgh 311 call analysis"
author: "Brian Bowling"
date: "August 23, 2018"
output: html_document
---

```{r libraries, echo=F, message=F,warning=F}

# general toolbox
library(tidyverse)
# for reading shapefiles, and simple features data frames
library(sf)
# for year and month functions
library(lubridate)
# for interactive mapping
library(leaflet)
# for better looking tables
library(pander)


```

```{r load data, echo=F, message=F, warning=F, results='hide'}

# downloaded raw data from https://data.wprdc.org/dataset/311-data and renamed file to
#   pitt311data.csv
data_311 <- read_csv('raw_data/pitt311data.csv')
# read in edited category file, original came from wPRDC
categories <- read_csv('raw_data/311-codebook-request-types-revised.csv')
# map files
block_map <- st_read('raw_data/pitt_bg_hood.shp')
# pop data
pop_data <- read_csv('raw_data/acs_16_5yr_b01003_with_ann.csv', col_types = 'cccii', skip=1)
# worker data (created with separate script, extract_pitt_lodes_data.R)
pitt_wac_bg <- read_csv('raw_data/pitt_wac.csv',col_types = cols(w_bg='c'))
pitt_rac_bg <- read_csv('raw_data/pitt_rac.csv',col_types = cols(h_bg='c'))


```

```{r data prep, echo=F, message=F}

# 311 data
# fix column names
colnames(data_311) = tolower(make.names(colnames(data_311)))

data_311 <- rename(data_311,'id'='x_id')

# add created_date field used in several analyses
data_311['created_date'] <- as.Date(data_311$created_on)

# eliminate months with partial data
# at the start
culls <- data_311$id[data_311$created_date < as.Date('2015-05-01')]
data_311 <- data_311[!data_311$id %in% culls,]
# at the end
culls <- data_311$id[data_311$created_date >= as.Date('2018-08-01')]
data_311 <- data_311[!data_311$id %in% culls,]

# get rid of calls dealing with issues outside of the city's boundaries
data_311 <- data_311[!data_311$geo_accuracy == 'OUT_OF_BOUNDS',]

# clean up and categorize request_type
data_311$request_type = tolower(data_311$request_type)

# trim categories to the columns needed in the join to the data_311 table
categories <- categories %>%
  select(request_type, category)

# join categories to data_311
data_311 <- left_join(data_311,categories, by='request_type')

# prepare for mapping
data_311 <- data_311 %>%
  rename(lon = x, lat = y)

# pitt block groups
pitt_bgs <- data.frame(bgs = unique(block_map$GEOID))



# population data prep
# fix column names and delete unneeded columns
colnames(pop_data) <- tolower(colnames(pop_data))
pop_data <- pop_data %>%
  select(geoid=id2, geography, pop=`estimate; total`)

# limit to pittsburgh block groups
pitt_pop_data <- pop_data %>%
  filter(pop_data$geoid %in% pitt_bgs$bgs)

```

```{r calculations, echo=F, message=F, warnings=F}

# calculate daily totals
dailies <- data_311 %>%
  group_by('year' = year(created_on), 'month' = month(created_on), 'day'=day(created_on)) %>%
  summarize(requests=n())

dailies_2015 <- dailies %>%
  filter(year == '2015')

dailies_2018 <- dailies %>%
  filter(year=='2018')

median_2015 = median(dailies_2015$requests)
median_2018 = median(dailies_2018$requests)
total = sum(dailies$requests)


```

```{r map calculations, echo=F, warning=F, message=F}

pitt_map <- left_join(block_map, pitt_pop_data, by=c('GEOID'='geoid'))

pitt_map <- left_join(pitt_map, pitt_wac_bg,by=c('GEOID'='w_bg'))

pitt_map <- left_join(pitt_map, pitt_rac_bg, by=c("GEOID"='h_bg'))

pitt_neighborhoods <- pitt_map %>%
  group_by(hood) %>%
  summarise(pop=sum(pop),
            workers_in_hood = sum(workers_in_bg,na.rm=T),
            workers_from_hood = sum(workers_from_bg,na.rm=T))


pitt_neighborhoods <- pitt_neighborhoods %>%
  mutate(daytime_pop = pop + (workers_in_hood - workers_from_hood))

neighborhood_monthlies <- data_311 %>%
  group_by(neighborhood, year = year(created_date), month = month(created_date)) %>%
  summarize(requests = n())

neighborhood_monthlies <- neighborhood_monthlies %>%
  filter(year==2018) %>%
  group_by(neighborhood) %>%
  summarize(monthly_median = median(requests))


neighborhood_request_count <- data_311 %>%
  group_by(neighborhood) %>%
  summarize(requests = n())

neighborhood_request_count <- left_join(neighborhood_request_count,neighborhood_monthlies,by='neighborhood')


data_311_map <- left_join(pitt_neighborhoods,neighborhood_request_count,by=c('hood'='neighborhood'))

data_311_map <- data_311_map %>%
  mutate(requests_per = round(requests/daytime_pop,digits=3)*100)





```




People contact Pittsburgh's 311 center to raise concerns about everything from cracked sidewalks to dangerous buildings.

About 42 percent of the 311 calls made in Pittsburgh concern road and street issues such as potholes, obstructions, etc. The next two largest categories are neighborhood issues such as derelict buildings and overgrown lots and garbage and litter issues.


Here's a list of the top 10 categories:
```{r, echo=F}
data_311 %>%
  group_by(category) %>%
  summarise(requests=n()) %>%
  arrange(desc(requests)) %>%
  mutate(pct_total = round(requests/sum(requests),digits=3)*100) %>%
  slice(1:10) %>%
  pander(big.mark=',',justify=c('left','right','right'),style='simple')


```

In terms of specific issues, potholes are the most frequent reason people contact the 311 Center. Here's a list of the top 10 issues:

```{r, echo=F}
data_311 %>%
  group_by(request_type) %>%
  summarize(requests = n()) %>%
  mutate(pct_total = round(requests/sum(requests),digits=3) * 100) %>%
  arrange(desc(requests)) %>%
  slice(1:10) %>%
  pander(big.mark=',',justify=c('left','right','right'),style='simple')

```


From May 2015 to July 2018, the 311 center has been contacted `r prettyNum(total,big.mark=',')` times about issues within the city's boundaries. The median number of 311 calls has gone from `r prettyNum(median_2015,big.mark=',')` per day in 2015 to `r prettyNum(median_2018,big.mark=',')` per day so far in 2018. 

The volume of requests follow a seasonal pattern with snow/ice removal driving up the 311 calls in winter months followed by an even higher number of pothole calls when things start thawing out.


```{r monthlies, echo=F}

data_311 %>%
  group_by('year' = year(created_on), 'month' = month(created_on)) %>%
  summarize(requests=n()) %>%
  ggplot(aes(x=paste0(year,'-',str_pad(month,2,pad='0')),y=requests)) +
  geom_col(color='blue',fill='blue',width=.8) +
  expand_limits(y=0) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0.5),
        plot.title = element_text(hjust=0.5),
        panel.grid = element_line(color='transparent')) +
  labs(title = 'Number of 311 Requests by Year and Month',
       x='Year-Month', y='Requests')

```


```{r snow potholes, echo=F}

data_311 %>%
  filter(request_type %in% c('potholes','snow/ice removal')) %>%
  group_by(year=year(created_on), month=month(created_on),request_type) %>%
  summarize(requests = n()) %>%
  ggplot(aes(x=paste0(year,'-',str_pad(month,2,pad='0')),y=requests)) +
  geom_col(color='blue',fill='blue',width=0.8) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0.5),
        plot.title = element_text(hjust=0.5)) +
  labs(title = 'Number of pothole and snow removal requests by Year and Month',
       x='Year-Month', y='Requests') +
  facet_wrap(~request_type,nrow=2)
```

# Maps
Here's a map showing the total requests and the 2018 median monthly requests by neighborhood.


```{r map it, echo=F, message=F, warning=F}
popup_sb <- paste0(data_311_map$hood,", Total = ",
                       as.character(prettyNum(data_311_map$requests,big.mark=',')),
                       ", Monthly median = ",
                       as.character(data_311_map$monthly_median))

pal_sb <- colorNumeric("Reds", domain=data_311_map$requests)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(-80, 40.44, zoom=12) %>%
  addPolygons(data=data_311_map,
              fillColor= ~pal_sb(data_311_map$requests),
              fillOpacity=0.9,
              weight=0.2,
              smoothFactor = 0.2,
              highlightOptions = highlightOptions(
                weight=5,
                color= "#666",
                fillOpacity = 0.7,
                bringToFront = TRUE
              ),
              label=popup_sb,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto")
              ) %>%
  addLegend(pal = pal_sb, 
            values = data_311_map$requests, 
            position = "bottomright", 
            title = "311 calls by Neighborhood")

```

Here's a map showing the total call per person using each neighborhood's estimated daytime population as the denominator. The daytime population was calculated using 2016 American Community Survey population estimates and Census Bureau estimates of the number of jobs and employed residents in each area. The basic formula is area population + (workers with jobs in the area - residents in the area who have jobs). This accounts for people who live outside the area but work in the area as well as residents who work outside of the area.

```{r per capita map, echo=F, warning=F, message=F}

popup_sb <- paste0(data_311_map$hood,", Per person = ",
                       as.character(data_311_map$requests_per))

pal_sb <- colorNumeric("Reds", domain=data_311_map$requests_per)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(-80, 40.44, zoom=12) %>%
  addPolygons(data=data_311_map,
              fillColor= ~pal_sb(data_311_map$requests_per),
              fillOpacity=0.9,
              weight=0.2,
              smoothFactor = 0.2,
              highlightOptions = highlightOptions(
                weight=5,
                color= "#666",
                fillOpacity = 0.7,
                bringToFront = TRUE
              ),
              label=popup_sb,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto")
              ) %>%
  addLegend(pal = pal_sb, 
            values = data_311_map$requests_per, 
            position = "bottomright", 
            title = "311 calls per daytime population")

```

